---
title: "DSMZ–TCGA Cohort Mapping by Organ and Spearman Similarity"
date: "2025-08-08"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  dev = "pdf", fig.ext = "pdf",
  fig.path = "/mnt/beegfs/mnt/work/bioinfo/home/chu25/dsmz/results/preliminary_analysis/"
)
```

```{r load-packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(SummarizedExperiment)
  library(pheatmap)
  library(matrixStats)  # ok to keep; not used for reduction here
  library(dplyr)
})

logmsg <- function(fmt, ...) message(sprintf(fmt, ...))
# Lightweight %||% (avoid rlang dependency)
`%||%` <- function(x, y) if (is.null(x)) y else x
```


# TCGA SummarizedExperiment (genes x samples)
# Load DSMZ and TCGA SummarizedExperiment 
```{r load-data}
# TCGA SummarizedExperiment (genes x samples)
tcga_se     <- readRDS("/home/chu25/data/tcga/ALL_TCGA_STAR_Counts_SummarizedExperiment.rds")
tcga_counts <- assay(tcga_se)  # numeric/integer matrix expected

# DSMZ counts table: Ensembl_ID, gene_name, then sample columns
dsmz_raw  <- readRDS("/home/chu25/data/dsmz/DSMZ_count_gene.rds")
dsmz_meta <- read.csv("/home/chu25/data/dsmz/DSMZ_metadata_with_sample_name.csv")

stopifnot(all(c("Ensembl_ID","gene_name") %in% colnames(dsmz_raw)))
logmsg("Loaded: TCGA (%d genes x %d samples) | DSMZ raw table: %d rows x %d cols",
       nrow(tcga_counts), ncol(tcga_counts), nrow(dsmz_raw), ncol(dsmz_raw))

```



# Align DSMZ metadata sample_name to columns in counts
# Keep only exact matches; report non-matching rows

```{r dsmz_counts_matrix}
# sample columns = everything except annotations
sample_cols <- setdiff(colnames(dsmz_raw),
                       c("Ensembl_ID","gene_name","Ensembl_ID_with_version"))

# force numeric (robust to misc formats)
dsmz_raw[sample_cols] <- lapply(dsmz_raw[sample_cols], function(x) as.numeric(as.character(x)))

# drop DSMZ columns that are all NA (after coercion)
keep <- vapply(as.data.frame(dsmz_raw[sample_cols]), function(v) any(!is.na(v)), logical(1))
if (!all(keep)) {
  logmsg("Dropping %d DSMZ sample columns that are all NA after coercion.", sum(!keep))
  sample_cols <- sample_cols[keep]
}

# matrix (genes x samples); rownames = Ensembl_ID (no version)
dsmz_counts <- as.matrix(dsmz_raw[, sample_cols, drop = FALSE])
rownames(dsmz_counts) <- dsmz_raw$Ensembl_ID

# if duplicated Ensembl_ID exist (multiple rows per gene), aggregate by sum
if (any(duplicated(rownames(dsmz_counts)))) {
  dsmz_counts <- rowsum(dsmz_counts, group = rownames(dsmz_counts), reorder = TRUE)
  logmsg("Collapsed duplicated Ensembl_ID rows by sum; now %d genes.", nrow(dsmz_counts))
}

logmsg("DSMZ counts dims: %d genes x %d samples", nrow(dsmz_counts), ncol(dsmz_counts))

```

```{r save_unmatched_2_csv}
stopifnot("sample_name" %in% names(dsmz_meta))
dsmz_meta <- dsmz_meta %>% mutate(sample_id = sample_name)

matched_ids   <- intersect(dsmz_meta$sample_id, colnames(dsmz_counts))
unmatched_md  <- setdiff(dsmz_meta$sample_id, colnames(dsmz_counts))
unmatched_cnt <- setdiff(colnames(dsmz_counts), dsmz_meta$sample_id)

logmsg("Matched DSMZ samples: %d | Unmatched (metadata): %d | Unmatched (counts): %d",
       length(matched_ids), length(unmatched_md), length(unmatched_cnt))

# save unmatched lists for audit
qc_dir <- "/home/chu25/dsmz/results/preliminary_analysis/matching_checks"
dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)
if (length(unmatched_md))
  write.csv(tibble(sample_name = unmatched_md),
            file.path(qc_dir, "unmatched_md.csv"), row.names = FALSE)
if (length(unmatched_cnt))
  write.csv(tibble(sample_name = unmatched_cnt),
            file.path(qc_dir, "unmatched_cnt.csv"), row.names = FALSE)

# keep only matched & reorder counts to metadata order
dsmz_meta   <- dsmz_meta %>% filter(sample_id %in% matched_ids)
dsmz_counts <- dsmz_counts[, dsmz_meta$sample_id, drop = FALSE]
stopifnot(identical(colnames(dsmz_counts), dsmz_meta$sample_id))

```

# Create organ labels from the DSMZ "Origin" field
```{r create-dsmz_meta-organ-column}
dsmz_meta <- dsmz_meta %>%
  mutate(
    organ = case_when(
      str_detect(Origin, regex("breast",  TRUE)) ~ "Breast",
      str_detect(Origin, regex("lung",    TRUE)) ~ "Lung",
      str_detect(Origin, regex("liver|hepato|bile", TRUE)) ~ "Liver/Biliary",
      str_detect(Origin, regex("adrenal", TRUE)) ~ "Adrenal",
      str_detect(Origin, regex("brain|gli", TRUE)) ~ "Brain/CNS",
      str_detect(Origin, regex("lymph|mantle|hodgkin|effusion|b[- ]?cell|t[- ]?cell|nk|natural\\s*killer", TRUE)) ~ "Lymphoid",
      str_detect(Origin, regex("leukemia", TRUE)) ~ "Leukemia",
      str_detect(Origin, regex("myeloid|megakaryocytic|erythroid|monocytic", TRUE)) ~ "Myeloid",
      str_detect(Origin, regex("bone\\s*marrow", TRUE)) ~ "Bone Marrow",
      str_detect(Origin, regex("abdomen|pleural", TRUE)) ~ "Body Cavity",
      is.na(Origin) | Origin == "NONE" ~ "Unknown",
      TRUE ~ "Other"
    )
  )
```

# Barplot to PDF
```{r Barplot_dsmz_metadata}
organ_counts <- dsmz_meta %>% dplyr::count(organ, name = "n_lines") %>% dplyr::arrange(desc(n_lines))
pdf("/home/chu25/dsmz/results/preliminary_analysis/DSMZ_organ_distribution.pdf", width = 6, height = 4)
ggplot(organ_counts, aes(x = reorder(organ, n_lines), y = n_lines, fill = organ)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = n_lines), vjust = -0.2, size = 3) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Number of DSMZ cell lines",
       title = "Organ distribution of DSMZ collection") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 40, hjust = 1))
dev.off()
```


# Harmonize genes (keep only shared Ensembl IDs)
```{r Harmonize-intersect_genes}
# numeric safety for correlations
storage.mode(dsmz_counts) <- "double"
storage.mode(tcga_counts) <- "double"

strip_ver <- function(x) sub("\\.\\d+$", "", x)

# TCGA: strip version and collapse duplicates (sum); keep ordered for reproducibility
tcga_base <- strip_ver(rownames(tcga_counts))
logmsg("TCGA before harmonize: %d genes", nrow(tcga_counts))
tcga_counts <- rowsum(tcga_counts, group = tcga_base, reorder = TRUE)
logmsg("TCGA after strip+collapse: %d unique base Ensembl genes", nrow(tcga_counts))


common_genes <- intersect(rownames(tcga_counts), rownames(dsmz_counts))

logmsg("head(rownames(tcga_counts)): %s",
       paste(head(rownames(tcga_counts)), collapse = ", "))

logmsg("head(rownames(dsmz_counts)): %s",
       paste(head(rownames(dsmz_counts)), collapse = ", "))
logmsg("Common genes (DSMZ ∩ TCGA): %d", length(common_genes))

# RELAXED: warn if small, but do NOT stop
if (length(common_genes) < 1000)
  logmsg("WARNING: Only %d shared genes. Correlations may be less stable.", length(common_genes))

tcga_counts <- tcga_counts[common_genes, , drop = FALSE]
dsmz_counts <- dsmz_counts[common_genes, , drop = FALSE]
stopifnot(identical(rownames(tcga_counts), rownames(dsmz_counts)))
```

```{r TCGA_cohort_labels}
# choose label column with fallback
tcga_labels <- NULL
cand <- c("project_id","study","disease_type")
for (nm in cand)
  if (nm %in% colnames(colData(tcga_se))) { tcga_labels <- colData(tcga_se)[[nm]]; break }
stopifnot(!is.null(tcga_labels))

if (is.factor(tcga_labels)) tcga_labels <- as.character(tcga_labels)
tcga_labels <- sub("^TCGA-", "", tcga_labels)
names(tcga_labels) <- colnames(tcga_counts)
stopifnot(length(tcga_labels) == ncol(tcga_counts))

tcga_levels <- sort(unique(tcga_labels))
tcga_means <- sapply(tcga_levels, function(ct) {
  rowMeans(tcga_counts[, tcga_labels == ct, drop = FALSE])
})
colnames(tcga_means) <- tcga_levels
stopifnot(identical(rownames(tcga_means), rownames(tcga_counts)))
logmsg("TCGA cohorts available: %d", length(tcga_levels))
```


#Organ → TCGA cohorts (broad)
```{r TCGA_Broad_organ}
organ2tcga <- list(
  "Adrenal"       = c("ACC","PCPG"),
  "Bladder"       = "BLCA",
  "Brain/CNS"     = c("GBM","LGG"),
  "Breast"        = "BRCA",
  "Cervix"        = "CESC",
  "Liver/Biliary" = c("LIHC","CHOL"),
  "Colon/Rectum"  = c("COAD","READ"),
  "Esophagus"     = "ESCA",
  "Head/Neck"     = "HNSC",
  "Kidney"        = c("KICH","KIRC","KIRP"),
  "Lung"          = c("LUAD","LUSC"),
  "Mesothelium"   = "MESO",
  "Ovary"         = "OV",
  "Pancreas"      = "PAAD",
  "Prostate"      = "PRAD",
  "Sarcoma"       = "SARC",
  "Skin"          = "SKCM",
  "Stomach"       = "STAD",
  "Testis"        = "TGCT",
  "Thyroid"       = "THCA",
  "Thymus"        = "THYM",
  "Uterus"        = c("UCEC","UCS"),
  "Uveal/Eye"     = "UVM",
  "Lymphoid"      = "DLBC",
  "Leukemia"      = "LAML",
  "Myeloid"       = "LAML",
  "Bone Marrow"   = "LAML",
  "Body Cavity"   = "MESO",
  "Other"         = character(0),
  "Unknown"       = character(0)
)
```

# Spearman similarity (DSMZ sample x TCGA cohort), using shared genes
```{r spearman_similarity}
g <- rownames(tcga_means)  # all shared genes
logmsg("Spearman will use %d genes across %d TCGA cohorts and %d DSMZ samples.",
       length(g), ncol(tcga_means), ncol(dsmz_counts))

spearman_scores <- outer(
  dsmz_meta$sample_id, colnames(tcga_means),
  Vectorize(function(cell, cohort) {
    cor(dsmz_counts[g, cell], tcga_means[g, cohort],
        method = "spearman", use = "pairwise.complete.obs")
  })
)
dimnames(spearman_scores) <- list(dsmz_meta$sample_id, colnames(tcga_means))
```


```{r TCGA_cohort_mapping}
# Make sure every organ label is one of the keys in organ2tcga
# Ensure every organ string is a key in organ2tcga (else "Unknown")
dsmz_meta <- dsmz_meta %>%
  mutate(organ = ifelse(organ %in% names(organ2tcga), organ, "Unknown"))

# 1) Build a list-column of allowed cohorts per row
dsmz_meta <- dsmz_meta %>%
  mutate(
    allowed = lapply(organ, function(org) {
      a <- organ2tcga[[org]]
      if (is.null(a)) a <- character(0)
      # keep only cohorts that actually exist in tcga_means
      intersect(a, colnames(tcga_means))
    })
  )

# 2) Pick the best cohort among the allowed list using Spearman
dsmz_meta <- dsmz_meta %>%
  mutate(
    tcga_code = mapply(function(sid, cand) {
      if (is.na(sid) || length(cand) == 0 || !(sid %in% rownames(spearman_scores))) {
        return(NA_character_)
      }
      v <- spearman_scores[sid, cand, drop = TRUE]
      if (length(v) == 0 || all(is.na(v))) NA_character_ else cand[which.max(v)]
    }, sample_id, allowed, USE.NAMES = FALSE)
  )

# quick diagnostics (optional)
message("Rows with no allowed cohorts: ", sum(lengths(dsmz_meta$allowed) == 0))


# sanity table
tab <- as.data.frame(
  table(Organ = dsmz_meta$organ, TCGA = dsmz_meta$tcga_code, useNA = "ifany")
) %>%
  arrange(Organ, TCGA)
utils::head(tab, 20)

```

```{r heatmap_organ_tcga_dsmz}
counts_long <- dsmz_meta %>%
  mutate(TCGA = ifelse(is.na(tcga_code), "Unassigned", tcga_code)) %>%
  dplyr::count(Organ = organ, TCGA, name = "Freq")

heat_mat <- counts_long %>%
  tidyr::pivot_wider(names_from = TCGA, values_from = Freq, values_fill = 0) %>%
  tibble::column_to_rownames("Organ") %>%
  as.matrix()

pdf("/home/chu25/dsmz/results/preliminary_analysis/organ_tcga_counts_heatmap.pdf", width = 8, height = 6)
pheatmap::pheatmap(
  heat_mat,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = TRUE, number_color = "black",
  main = "DSMZ organ → TCGA cohort assignments"
)
dev.off()
```

```{r save_spearman_outputs}
out_dir <- "/home/chu25/dsmz/results/preliminary_analysis/spearman"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# all Spearman (wide & long) + RDS
all_wide <- spearman_scores %>%
  as.data.frame() %>% tibble::rownames_to_column("sample_id")
write.csv(all_wide, file.path(out_dir, "spearman_scores_all_wide.csv"), row.names = FALSE)

all_long <- tidyr::pivot_longer(all_wide, -sample_id,
                                names_to = "tcga_cohort", values_to = "rho")
write.csv(all_long, file.path(out_dir, "spearman_scores_all_long.csv"), row.names = FALSE)
saveRDS(spearman_scores, file.path(out_dir, "spearman_scores_all.rds"))

# max overall per DSMZ sample (ignoring organ)
best_idx <- max.col(spearman_scores, ties.method = "first")
best_overall_df <- tibble::tibble(
  sample_id         = rownames(spearman_scores),
  best_tcga_overall = colnames(spearman_scores)[best_idx],
  best_rho_overall  = spearman_scores[cbind(seq_len(nrow(spearman_scores)), best_idx)]
)

# rho for organ-constrained assignment (if any)
rho_allowed <- rep(NA_real_, nrow(dsmz_meta))
ok <- !is.na(dsmz_meta$tcga_code) & dsmz_meta$sample_id %in% rownames(spearman_scores)
rho_allowed[ok] <- spearman_scores[cbind(dsmz_meta$sample_id[ok], dsmz_meta$tcga_code[ok])]

max_summary <- dsmz_meta %>%
  dplyr::select(sample_id, Origin, organ, tcga_code) %>%
  dplyr::left_join(best_overall_df, by = "sample_id") %>%
  dplyr::mutate(rho_allowed = rho_allowed)

write.csv(max_summary, file.path(out_dir, "spearman_max_by_sample.csv"), row.names = FALSE)

# keep the mapping as a simple table too
write.csv(dsmz_meta %>% dplyr::select(sample_id, Origin, organ, tcga_code),
          file.path(out_dir, "dsmz_tcga_organ_mapping.csv"), row.names = FALSE)
```


